---
title: "Bronch Mukami"
author: " "
date: "2023-07-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## CEMA Internship Task
You have been provided with a dataset (in the data folder above) which contains monthly data for children <5 years, disaggregated at a county level for the period January 2021 to June 2023.

### The dataset contains the following variables:

1) Period (months from January 2021 to June 2023)
2) County (the 47 counties in Kenya)
3) Total number of children dewormed (Total Dewormed)
4) Number of children <5 years with acute malnutrition (Acute Malnutrition)
5) Number of children stunted (0-6 months, 6-23 months, 24-59 months)
6) Number of children <5 years with diarrhea (Diarrhea cases)
7) Number of children who are underweight (0-6 months, 6-23 months, 24-59 months)

### Your task is to:

1) Conduct exploratory data analysis
2) State an appropriate research question you would want to answer from the data
3) Carry out appropriate data analysis to address the question you have stated above

### Research Question
Is there a relationship between the prevalence of acute malnutrition and other indices of stunting and underweight across various age groups in various counties over time?

```{r}
#import the libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r}
#loading the dataset
taskdf= read.csv("https://raw.githubusercontent.com/cema-uonbi/internship_task/main/data/cema_internship_task_2023.csv")
head(taskdf)
```

### 1) Data Preprocessing

```{r}
#to get the data structure 
str(taskdf)
```
```{r}
#to get a data summary
summary(taskdf)
```
```{r}
#to check the shape of the data
dim(taskdf)
```

```{r}
#to check for null values
null_values <- is.na(taskdf)

#to count the number of null values in each column
col_null_counts <- colSums(null_values)

#to display the count of null values for each column
print(col_null_counts)
```
```{r}
#to remove rows with null values
taskdf_clean <- taskdf[complete.cases(taskdf),]

#to check the number of rows in the cleaned data frame
nrow(taskdf_clean)
```
```{r}
head(taskdf_clean)
```
```{r}
#convert all column names to lower case
names(taskdf_clean) <- tolower(names(taskdf_clean))
head(taskdf_clean)
```

```{r}
names(taskdf_clean)
```

```{r}
#to rename some columns
taskdf_clean <- taskdf_clean %>%
  rename(
    `total-dewormed` = total.dewormed,
    `acute-malnutrition` = acute.malnutrition,
    `stunted6-23months` = stunted.6.23.months,
    `stunted0-6months` = stunted.0..6.months,
    `stunted24-59months` = stunted.24.59.months,
    `diarrhoea-cases` = diarrhoea.cases,
    `underweight0-6months` = underweight.0..6.months,
    `underweight6-23months` = underweight.6.23.months,
    `underweight24-59months` = underweight.24.59.months
  )
names(taskdf_clean)
```

### 2) Exploratory Data Analysis
```{r}
#to get unique values of county column
unique_counties <- unique(taskdf_clean$county)
print(unique_counties)

#to get the number of unique values in county column
num_unique_counties <- length(unique_counties)
print(num_unique_counties)
```
#### * This shows that there is complete data from 44 out of 47 counties in Kenya

```{r}
library(dplyr)

#convert 'period' column to Date class with the correct year
taskdf_clean$period <- as.Date(paste0("01-", taskdf_clean$period), format = "%d-%b-%y")

#group by county and year, and summarize to find maximum and minimum values
max_min_summary <- taskdf_clean %>%
  group_by(county, year = format(period, "%Y")) %>%
  summarize(
    max_dewormed = max(`total-dewormed`, na.rm = TRUE),
    min_dewormed = min(`total-dewormed`, na.rm = TRUE),
    max_acute_malnutrition = max(`acute-malnutrition`, na.rm = TRUE),
    min_acute_malnutrition = min(`acute-malnutrition`, na.rm = TRUE),
    max_stunted6_23months = max(`stunted6-23months`, na.rm = TRUE),
    min_stunted6_23months = min(`stunted6-23months`, na.rm = TRUE),
    max_stunted0_6months = max(`stunted0-6months`, na.rm = TRUE),
    min_stunted0_6months = min(`stunted0-6months`, na.rm = TRUE),
    max_stunted24_59months = max(`stunted24-59months`, na.rm = TRUE),
    min_stunted24_59months = min(`stunted24-59months`, na.rm = TRUE),
    max_underweight0_6months = max(`underweight0-6months`, na.rm = TRUE),
    min_underweight0_6months = min(`underweight0-6months`, na.rm = TRUE),
    max_underweight6_23months = max(`underweight6-23months`, na.rm = TRUE),
    min_underweight6_23months = min(`underweight6-23months`, na.rm = TRUE),
    max_underweight24_59months = max(`underweight24-59months`, na.rm = TRUE),
    min_underweight24_59months = min(`underweight24-59months`, na.rm = TRUE)
  )

max_min_summary

```

#### 2.1 Correlation of Variables
```{r}
#select numeric columns for correlation matrix calculation
numeric_columns <- taskdf_clean[, sapply(taskdf_clean, is.numeric)]

#calculate the correlation matrix
correlation_matrix <- cor(numeric_columns, use = "complete.obs")
print(correlation_matrix)

```

```{r}
#plot correlation matrix as a heatmap
install.packages("corrplot")
library(corrplot)

corrplot(correlation_matrix, method = "color", tl.cex = 0.7, tl.col = "black")
```

#### 2.2 Visualizations
```{r}
#creating scatter plots
library(ggplot2)
ggplot(taskdf_clean, aes(x = `total-dewormed`, y = `acute-malnutrition`, color= "red")) +
  geom_point() +
  labs(x = "Total Dewormed", y = "Acute Malnutrition") +
  ggtitle("Scatter Plot: Total Dewormed vs. Acute Malnutrition")

```
```{r}
#create scatter plots for acute malnutrition vs. stunting indicators
ggplot(taskdf_clean, aes(x = `acute-malnutrition`, y = `stunted0-6months`, color= "blue")) +
  geom_point() +
  labs(x = "Prevalence of Acute Malnutrition", y = "Stunted 0-6 months") +
  ggtitle("Scatter Plot: Acute Malnutrition vs. Stunted 0-6 months")

ggplot(taskdf_clean, aes(x = `acute-malnutrition`, y = `stunted6-23months`, color="orange")) +
  geom_point() +
  labs(x = "Prevalence of Acute Malnutrition", y = "Stunted 6-23 months") +
  ggtitle("Scatter Plot: Acute Malnutrition vs. Stunted 6-23 months")

ggplot(taskdf_clean, aes(x = `acute-malnutrition`, y = `stunted24-59months`, color="magenta")) +
  geom_point() +
  labs(x = "Prevalence of Acute Malnutrition", y = "Stunted 24-59 months") +
  ggtitle("Scatter Plot: Acute Malnutrition vs. Stunted 24-59 months")

```
```{r}
ggplot(taskdf_clean, aes(x = period, y = `acute-malnutrition`, color="blue")) +
  geom_area() +
  labs(x = "Period", y = "Acute Malnutrition") +
  ggtitle("Acute Malnutrition over Time") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```


```{r}
install.packages("RColorBrewer")
library(RColorBrewer)
```

```{r}
#filter the data to include only the relevant stunting and underweight indicators
stunting_indicators <- c("stunted0-6months", "stunted6-23months", "stunted24-59months")
underweight_indicators <- c("underweight0-6months", "underweight6-23months", "underweight24-59months")

stunting_df <- taskdf_clean[, c("period", stunting_indicators)]
underweight_df <- taskdf_clean[, c("period", underweight_indicators)]

#melt the data to long format for line plotting
install.packages("reshape2")
library(reshape2)

stunting_melted <- melt(stunting_df, id.vars = "period")
underweight_melted <- melt(underweight_df, id.vars = "period")

#create custom color palettes for stunting and underweight indicators
library(RColorBrewer)

stunting_color_palette <- brewer.pal(n = length(stunting_indicators), name = "Set2")
underweight_color_palette <- brewer.pal(n = length(underweight_indicators), name = "Paired")

#create area plots for stunting indicators
ggplot(stunting_melted, aes(x = period, y = value, fill = variable)) +
  geom_area() +
  scale_fill_manual(values = stunting_color_palette) +  # Use stunting color palette
  labs(x = "Period", y = "Prevalence", fill = "Stunting Indicator") +
  ggtitle("Time Series of Stunting Indicators") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#create area plots for underweight indicators
ggplot(underweight_melted, aes(x = period, y = value, fill = variable)) +
  geom_area() +
  scale_fill_manual(values = underweight_color_palette) +  # Use underweight color palette
  labs(x = "Period", y = "Prevalence", fill = "Underweight Indicator") +
  ggtitle("Time Series of Underweight Indicators") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### 3) Results and Conclusion
* There is a Positive Correlation between the Underweight Indicators and the Stunting Indicators respectively per their ages i.e children who were underweight(0-6)months may also have been stunted at (0-6)months.

* The number of children who were stunted increased over the years with the year 2023 being the peak. It is also noticeable that children who were stunted at (6-23) months were the most while those at (24-59)months were the least.

* The number of children who were underweight increased over the years with the year 2023 being the peak. t is also noticeable that children who were stunted at (6-23) months were the most while those at (24-59)months were the least.